---
title: "Practical Machine Learning Final Project"
author: "JADraves"
date: "September 28, 2019"
output: html_document
---

```{r setup, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret); library(ggplot2); library(reshape2); library(dplyr); library(gbm)
```

##Background 
###Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Extracting and Loading Data
###Links to the data were provide and .csv files were downloaded to local system
```{r Extracting and loading data, eval=FALSE}
train <- read.csv("pml-training.csv", sep = ",")
QuizData <-  read.csv("pml-testing.csv", sep = ",")
```


##Investigating Data Sets
###There are a great many observations with a value of "NA" and these will not be useful so thye must be removed.
```{r Investigating Data Sets, eval=FALSE}
##The following commands were used to preliminarily investigate the data, data types, and levels.
### head(train), str(train), tail(train), summary(train). Their output is not included due to length.
#The size of the data set and number of NA's are determined as follows:

dim(train)
## 19622   160

table(is.na(train))
##   FALSE    TRUE 
##  1852048 1287472 
```

###Some 1,287,472 values of "NA" were identified.  Further, many variables contain "NA's" for all observations. 
###All variables with a greater than 90% of their observations equivalent to "NA" are removed and a new data frame called CleanTrain is produced.  Also, the first 7 columns are only desciptive data so they are also removed.

```{r Producing a Clean Data Set, eval=FALSE}
### removing NA's to produce a Clean Trained Data Set

colrmv <- which(colSums(is.na(train) |train=="")>0.9*dim(train)[1])
CleanTrain <- train[,-colrmv]
CleanTrain <- CleanTrain[,-c(1:7)]

dim(CleanTrain)
## 19622    53

#CleanTrain contains the following:
str(CleanTrain)

##'data.frame':	19622 obs. of  53 variables:
## roll_belt           : num  1.41 1.41 1.42 1.48 1.48 1.45 1.42 1.42 1.43 1.45 ...
## pitch_belt          : num  8.07 8.07 8.07 8.05 8.07 8.06 8.09 8.13 8.16 8.17 ...
## yaw_belt            : num  -94.4 -94.4 -94.4 -94.4 -94.4 -94.4 -94.4 -94.4 -94.4 -94.4 ...
## total_accel_belt    : int  3 3 3 3 3 3 3 3 3 3 ...
## gyros_belt_x        : num  0 0.02 0 0.02 0.02 0.02 0.02 0.02 0.02 0.03 ...
## gyros_belt_y        : num  0 0 0 0 0.02 0 0 0 0 0 ...
## gyros_belt_z        : num  -0.02 -0.02 -0.02 -0.03 -0.02 -0.02 -0.02 -0.02 -0.02 0 ...
## accel_belt_x        : int  -21 -22 -20 -22 -21 -21 -22 -22 -20 -21 ...
## accel_belt_y        : int  4 4 5 3 2 4 3 4 2 4 ...
## accel_belt_z        : int  22 22 23 21 24 21 21 21 24 22 ...
## magnet_belt_x       : int  -3 -7 -2 -6 -6 0 -4 -2 1 -3 ...
## magnet_belt_y       : int  599 608 600 604 600 603 599 603 602 609 ...
## magnet_belt_z       : int  -313 -311 -305 -310 -302 -312 -311 -313 -312 -308 ...
## roll_arm            : num  -128 -128 -128 -128 -128 -128 -128 -128 -128 -128 ...
## pitch_arm           : num  22.5 22.5 22.5 22.1 22.1 22 21.9 21.8 21.7 21.6 ...
## yaw_arm             : num  -161 -161 -161 -161 -161 -161 -161 -161 -161 -161 ...
## total_accel_arm     : int  34 34 34 34 34 34 34 34 34 34 ...
## gyros_arm_x         : num  0 0.02 0.02 0.02 0 0.02 0 0.02 0.02 0.02 ...
## gyros_arm_y         : num  0 -0.02 -0.02 -0.03 -0.03 -0.03 -0.03 -0.02 -0.03 -0.03 ...
## gyros_arm_z         : num  -0.02 -0.02 -0.02 0.02 0 0 0 0 -0.02 -0.02 ...
## accel_arm_x         : int  -288 -290 -289 -289 -289 -289 -289 -289 -288 -288 ...
## accel_arm_y         : int  109 110 110 111 111 111 111 111 109 110 ...
## accel_arm_z         : int  -123 -125 -126 -123 -123 -122 -125 -124 -122 -124 ...
## magnet_arm_x        : int  -368 -369 -368 -372 -374 -369 -373 -372 -369 -376 ...
## magnet_arm_y        : int  337 337 344 344 337 342 336 338 341 334 ...
## magnet_arm_z        : int  516 513 513 512 506 513 509 510 518 516 ...
## roll_dumbbell       : num  13.1 13.1 12.9 13.4 13.4 ...
## pitch_dumbbell      : num  -70.5 -70.6 -70.3 -70.4 -70.4 ...
## yaw_dumbbell        : num  -84.9 -84.7 -85.1 -84.9 -84.9 ...
## total_accel_dumbbell: int  37 37 37 37 37 37 37 37 37 37 ...
## gyros_dumbbell_x    : num  0 0 0 0 0 0 0 0 0 0 ...
## gyros_dumbbell_y    : num  -0.02 -0.02 -0.02 -0.02 -0.02 -0.02 -0.02 -0.02 -0.02 -0.02 ...
## gyros_dumbbell_z    : num  0 0 0 -0.02 0 0 0 0 0 0 ...
## accel_dumbbell_x    : int  -234 -233 -232 -232 -233 -234 -232 -234 -232 -235 ...
## accel_dumbbell_y    : int  47 47 46 48 48 48 47 46 47 48 ...
## accel_dumbbell_z    : int  -271 -269 -270 -269 -270 -269 -270 -272 -269 -270 ...
## magnet_dumbbell_x   : int  -559 -555 -561 -552 -554 -558 -551 -555 -549 -558 ...
## magnet_dumbbell_y   : int  293 296 298 303 292 294 295 300 292 291 ...
## magnet_dumbbell_z   : num  -65 -64 -63 -60 -68 -66 -70 -74 -65 -69 ...
## roll_forearm        : num  28.4 28.3 28.3 28.1 28 27.9 27.9 27.8 27.7 27.7 ...
## pitch_forearm       : num  -63.9 -63.9 -63.9 -63.9 -63.9 -63.9 -63.9 -63.8 -63.8 -63.8 ...
## yaw_forearm         : num  -153 -153 -152 -152 -152 -152 -152 -152 -152 -152 ...
## total_accel_forearm : int  36 36 36 36 36 36 36 36 36 36 ...
## gyros_forearm_x     : num  0.03 0.02 0.03 0.02 0.02 0.02 0.02 0.02 0.03 0.02 ...
## gyros_forearm_y     : num  0 0 -0.02 -0.02 0 -0.02 0 -0.02 0 0 ...
## gyros_forearm_z     : num  -0.02 -0.02 0 0 -0.02 -0.03 -0.02 0 -0.02 -0.02 ...
## accel_forearm_x     : int  192 192 196 189 189 193 195 193 193 190 ...
## accel_forearm_y     : int  203 203 204 206 206 203 205 205 204 205 ...
## accel_forearm_z     : int  -215 -216 -213 -214 -214 -215 -215 -213 -214 -215 ...
## magnet_forearm_x    : int  -17 -18 -18 -16 -17 -9 -18 -9 -16 -22 ...
## magnet_forearm_y    : num  654 661 658 658 655 660 659 660 653 656 ...
## magnet_forearm_z    : num  476 473 469 469 473 478 470 474 476 473 ...
## classe              : Factor w/ 5 levels "A","B","C","D",..: 1 1 1 1 1 1 1 1 1 1 ...
```
### This has substantially reduced the number of observations from 160 to 53 and has produced a more useable and manageable data set.

##Establishing a Training and Testing Set
###The sets are produced by assigning 70% of the data to the Training set and 30% of the Data to the Testing set.

```{r Establishing Training and Testing Data Sets, eval=FALSE}
#Establishig Training and Testing Data Sets...use a 70/30 split

inTrain = createDataPartition(y=CleanTrain$classe, p=0.7, list=FALSE)
training = CleanTrain[inTrain,]
testing = CleanTrain[-inTrain,]
```

##Producing the Models
###The models chosen for evaluation are the Random Forest and Gradient Boosting. 
```{r Models, eval=FALSE}
# Crossvalidation parameters are in ControlofFit
ControlofFit <- trainControl(method='cv', number=3, allowParallel = TRUE)

#Random Forest
RandFmod <- train(classe ~., method="rf", data=training, trControl=ControlofFit)
RandFmod

## Random Forest 

## 13737 samples
##   52 predictor
##    5 classes: 'A', 'B', 'C', 'D', 'E' 

## No pre-processing
## Resampling: Cross-Validated (3 fold) 
## Summary of sample sizes: 9159, 9157, 9158 
## Resampling results across tuning parameters:

## mtry  Accuracy   Kappa    
##   2    0.9885708  0.9855388
##  27    0.9880614  0.9848944
##  52    0.9842036  0.9800131

## Accuracy was used to select the optimal model using the largest value.
## The final value used for the model was mtry = 2.

#Gradient Boost
GradBmod <- train(classe ~., method="gbm", data=training, trControl=ControlofFit, verbose=FALSE)
GradBmod

## Stochastic Gradient Boosting 

## 13737 samples
##   52 predictor
##    5 classes: 'A', 'B', 'C', 'D', 'E' 

## No pre-processing
## Resampling: Cross-Validated (3 fold) 
## Summary of sample sizes: 9159, 9157, 9158 
## Resampling results across tuning parameters:

##  interaction.depth  n.trees  Accuracy   Kappa    
##  1                   50      0.7522746  0.6859794
##  1                  100      0.8209212  0.7733283
##  1                  150      0.8504767  0.8108117
##  2                   50      0.8549902  0.8162950
##  2                  100      0.9040548  0.8785838
##  2                  150      0.9293880  0.9106701
##  3                   50      0.8946644  0.8665963
##  3                  100      0.9403073  0.9244690
##  3                  150      0.9586522  0.9476912

## Tuning parameter 'shrinkage' was held constant at a value of 0.1
## Tuning parameter
## 'n.minobsinnode' was held constant at a value of 10
## Accuracy was used to select the optimal model using the largest value.
## The final values used for the model were n.trees = 150, interaction.depth = 3, shrinkage = 0.1
## and n.minobsinnode = 10.
```

##Crossvalidation using Testing set

```{r CrossValidation, eval=FALSE}
# Crossvalidation
predRandF <- predict(RandFmod, newdata = testing)
ConMRandF <- confusionMatrix(predRandF, testing$classe)

predGradBmod <- predict(GradBmod, newdata = testing)
ConMGradB <- confusionMatrix(predGradBmod, testing$classe)

print(paste("The Accuracy of the Random Forest Model is", round(ConMRandF$overall[1], 4)), quote=FALSE)
print(paste("The Out of Sample Error of the Random Forest Model is", round(1-ConMRandF$overall[1], 4)), quote=FALSE)

## The Accuracy of the Random Forest Model is 0.9907
## The Out of Sample Error of the Random Forest Model is 0.0093

print(paste("The Accuracy of the Gradient Boost Model is", round(ConMGradB$overall[1], 4)), quote=FALSE)
print(paste("The Out of Sample Error of the Gradient Boost Model is", round(1-ConMGradB$overall[1], 4)), quote=FALSE)

##The Accuracy of the Gradient Boost Model is 0.9563
## The Out of Sample Error of the Gradient Boost Model is 0.0437
```


##Validation and Quiz Data
### Given that the Accuracy of the Random Forest Model is better, it will be used to complete the validation and the quiz.

```{r Validation and Quiz, eval=FALSE}
predictQuiz <- predict(rfmod, newdata = QuizData)
predictQuiz

## B A B A A E D B A A B C B A E E A B B B
## Levels: A B C D E
```

## These values yielded a 20/20 on the quiz.
